---
title: "Getting started with reiof"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with reiof}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(reiof)
```

## What EIOF is

Intake's **Environmental Impacts of Foods (EIOF)** database gives country-specific
environmental impact values for more than 6,500 foods across seven impact areas
spanning five domains of planetary health. What makes it unusual is not only its
coverage but its adjustment factors: non-edible portion, processing and cooking
yield, so that the *form of food consumed* in a dietary recall can be matched
correctly to the *form of food* the impact data describes.

Alongside the database, Intake published a set of **dietary environmental impact
metrics** for use with food consumption survey data. These come in three
families:

| Family | Question it answers | Source |
|---|---|---|
| Absolute threshold | Is this diet below the per capita daily share of a planetary boundary? | Table 3 |
| Relative benchmark | Where does this diet sit in a global reference distribution? | Tables 4a–4g |
| Composite | Is this diet both healthy and environmentally sound? | Tables 5a, 5b |

All of them are age- and sex-specific, because average dietary energy
requirement differs across groups.

## What reiof is — and is not

`reiof` implements those metrics in R.

**It does not contain the EIOF database.** It ships the published *cut-offs* and
applies them to impact values you have already derived. The work of matching
food items to impact factors, applying the adjustment factors, and summing to
the respondent-day happens upstream; `reiof` starts where that ends. It also
does not calculate the Global Diet Quality Score — the composite metrics take a
GDQS value as an input.

So the input `reiof` expects is one row per respondent-day, with the impact
totals already summed:

```{r}
head(eiof_demo, 3)
```

The package has **no dependencies at all**. The practical consequence is a plain
vector API: you pass `df$biodiv`, not `biodiv`.

## The seven indicators, and their units

Units matter more here than in most analyses, because a unit error produces a
confidently wrong answer rather than an obvious one.

```{r}
eiof_list_indicators()
```

## Step 1 — the demographic group

Every cut-off is scaled to one of eight age/sex groups. This variable is
required by everything else.

```{r}
grp <- eiof_age_sex_group(age = eiof_demo$age, sex = eiof_demo$sex)
table(factor(grp, 1:8, attr(grp, "labels")))
```

### Which type should `sex` be?

**A factor is the recommended input**, and `eiof_demo$sex` is one. The levels
carry the meaning, so the data documents itself, `table()` prints "male" and
"female" rather than 1 and 2, and `reiof` can resolve it without being told
anything further.

Character works identically, and is what `read.csv()` gives you by default:

```{r}
eiof_age_sex_group(age = c(1.5, 40), sex = c("M", "F"))
```

**Numeric sex is deliberately refused** unless you say which code means male:

```{r, error = TRUE}
eiof_age_sex_group(age = c(1.5, 40), sex = c(1, 0))
```

This is not a missing feature. Survey data uses `1/2` and `0/1` about equally
often, and the two are indistinguishable from the values alone — both are just
two distinct numbers. If `reiof` guessed and guessed wrong, every result would
be silently inverted: no error, nothing visibly odd, every male counted as
female. Rather than carry that risk, it stops and asks.

Say the codes and it works:

```{r}
eiof_age_sex_group(age = c(1.5, 40), sex = c(1, 0), male = 1, female = 0)
```

But the better habit is to convert once, near the top of your script, after
which every later call is unambiguous:

```{r}
sex_f <- eiof_as_sex_factor(c(1, 0, 1, 0), male = 1, female = 0)
sex_f
```

### A note on the age boundaries

The paper prints the groups as "0-1.9", "2-4.9", "5-9.9" and "10+". `reiof`
reads these as half-open intervals: `age < 2`, `2 <= age < 5`, `5 <= age < 10`,
`age >= 10`. So a child aged 1.95 years is in the first group:

```{r}
eiof_age_sex_group(c(1.9, 1.95, 1.999, 2.0), rep("male", 4))
```

Testing `age <= 1.9` instead would leave everyone between 1.9 and 2.0
unassigned, and they would then vanish from every metric with no error raised.

## Step 2 — absolute threshold metrics (Table 3)

```{r}
ghg_ok <- eiof_absolute_threshold(eiof_demo$ghg, grp, "ghg")
table(ghg_ok, useNA = "ifany")
```

Read as: *"x% of one-day diets fall below the absolute per capita daily
threshold established for greenhouse gas emissions."*

The comparison is strict — the published cut-offs carry a `<` sign, so a diet
sitting exactly on the threshold is coded `0`, not `1`.

All seven at once:

```{r}
for (ind in eiof_indicators$indicator) {
  x <- eiof_absolute_threshold(eiof_demo[[ind]], grp, ind)
  cat(sprintf("%-12s %5.1f%% below threshold\n", ind, 100 * mean(x, na.rm = TRUE)))
}
```

To see which number a respondent was compared against:

```{r}
eiof_lookup_cutoff("ghg", metric = "abs")
```

## Step 3 — relative benchmark metrics (Tables 4a–4g)

```{r}
band <- eiof_relative_band(eiof_demo$biodiv, grp, "biodiv")
table(band, useNA = "ifany")
```

**This does not rank your respondents against each other.** The bands come from
a fixed external reference distribution, the same for every analysis anywhere in
the world. See `vignette("relative-benchmarks")` for what that distribution is,
how it was built, and why `quantile()` is the wrong tool.

## Step 4 — the composite metrics (Tables 5a, 5b)

Both combine a diet quality criterion (GDQS ≥ 23) with an environmental one.
They differ only in the environmental criterion: HESD uses the **absolute**
biodiversity threshold, HDLEI the **relative** 30th percentile.

```{r}
table(eiof_hesd(eiof_demo$biodiv, eiof_demo$gdqs, grp))
```

Each criterion is also available on its own, which is often what you want for
reporting:

```{r}
parts <- eiof_hesd(eiof_demo$biodiv, eiof_demo$gdqs, grp, form = "components")
head(parts, 4)
```

`healthy` and `env` are the two sub-indicators; `category` is the four-way
Table 5a classification; `binary` is category 1 against everything else. The
sub-indicators are produced by calling `eiof_healthy_diet()` and
`eiof_sustainable_diet()` internally, so they cannot drift from the standalone
functions.

Note the message about respondents under 2 years. The methods paper restricts
the composites to ages 2 and over, because the GDQS has not been validated below
that age — Table 5a prints six demographic columns, not eight. If your sample
includes under-twos, decide explicitly whether they belong in the denominator,
because it changes every percentage you report.

## Three things to know before reporting any of this

### 1. Eutrophication units are unresolved

The paper states eutrophication potential is in kg PO₄³⁻-eq, and the Table 3
absolute threshold is consistent with that (0.0077–0.0267). The Table 4f
relative benchmarks for the same quantity are 5.3–48.7 — about a thousand times
larger.

Read as grams, Table 4f places the absolute threshold between the 40th and 50th
percentile, exactly where the two land thresholds sit. Read as kilograms, every
diet on earth falls below the 10th percentile, which is not a coherent reading
of a percentile distribution. `reiof` therefore applies a ÷1000 rescaling to
relative eutrophication only.

**This is an inference, not a documented conversion.** It is isolated in the
`scale_to_db_unit` column of `eiof_cutoffs` so it can be switched off in one
place. Confirm with Intake before publishing any eutrophication result.

### 2. A low relative band is not evidence of sustainability

The reference distribution describes diets as they are, not as they would need
to be. Compare where each absolute threshold falls within it:

```{r}
for (ind in eiof_indicators$indicator) {
  thr <- unname(eiof_lookup_cutoff(ind, 4, "abs"))
  p   <- vapply(seq(10, 90, 10),
                function(q) unname(eiof_lookup_cutoff(ind, 4, "rel", q)),
                numeric(1))
  k <- sum(thr >= p)
  pos <- if (k == 0) "below P10" else if (k == 9) "at/above P90" else
    sprintf("P%d-P%d", k * 10, (k + 1) * 10)
  cat(sprintf("%-12s %s\n", ind, pos))
}
```

For greenhouse gases the planetary-boundary threshold lies **below the 10th
percentile**: a diet in band 1, the lowest band, still exceeds the boundary. For
water it lies **above the 90th**, so essentially every diet passes. The two
metric families answer different questions and will not agree. Report them
together; never use one as a proxy for the other.

### 3. HDLEI is nested inside HESD

```{r}
h <- eiof_hesd(eiof_demo$biodiv, eiof_demo$gdqs, grp, form = "binary")
d <- eiof_hdlei(eiof_demo$biodiv, eiof_demo$gdqs, grp, form = "binary")
table(HESD = h, HDLEI = d, useNA = "ifany")
```

The empty cell is structural, not a quirk of this sample. For biodiversity the
relative P30 cut-off sits below the absolute threshold in every demographic
group, by the same constant factor:

```{r}
vapply(1:8, function(g) {
  unname(eiof_lookup_cutoff("biodiv", g, "abs")) /
    unname(eiof_lookup_cutoff("biodiv", g, "rel", 30))
}, numeric(1))
```

Both are the same all-population figure scaled by the same energy-requirement
weight, so the ratio is constant by construction. Every HDLEI diet is therefore
also a HESD diet, and HESD prevalence can never fall below HDLEI prevalence.
Reporting both is still informative — the gap between them is the share of diets
sitting between the two cut-offs — but they are not two independent dimensions
and must not be presented as a 2×2.

## Before you compute anything: check the data

Everything above assumes your impact values are the right type, the right unit
and the right precision. They very often are not, and the failures are quiet.

```{r}
report <- eiof_check_data(
  eiof_demo,
  impacts = c(biodiv = "biodiv", ghg = "ghg"),
  age = "age", sex = "sex", gdqs = "gdqs"
)
summary(report)
```

`vignette("troubleshooting")` walks through a deliberately broken dataset, the
errors it produces, and how to fix each one.

## Survey weights

`reiof` returns respondent-level indicator vectors and nothing else. It applies
no survey design. Every prevalence you report should come from a design-aware
estimator — the **survey** package, or whatever your analysis uses — or both
your point estimates and your standard errors will be wrong for a complex
design.
