---
title: "Troubleshooting: fixing a broken dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting: fixing a broken dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(reiof)
```

This vignette works through a dataset with four real faults in it — the kinds
that arrive from a spreadsheet, a database export or another statistical
package. We run the checks, read the errors, fix each one, and re-run.

The point is not the fixes themselves. It is that **all four faults produce
plausible-looking numbers rather than obvious failures** if you skip the checks.

## The data

A copy of `eiof_demo` ships with the package, broken on purpose:

```{r}
path <- system.file("extdata", "eiof_broken.csv", package = "reiof")
broken <- read.csv(path, stringsAsFactors = FALSE)
str(broken[, c("age", "sex", "ghg", "water", "biodiv", "gdqs")])
```

Nothing there looks alarming. `biodiv` is a character column, but that is easy
to miss in a `str()` dump of eleven variables, and `head()` would show values
that read like numbers.

## Run the checks first

```{r}
report <- eiof_check_data(
  broken,
  impacts = c(biodiv = "biodiv", ghg = "ghg", water = "water"),
  age = "age", sex = "sex", gdqs = "gdqs"
)
report
```

Four problems, found in one pass. Taking them in turn.

## Fault 1 — biodiversity imported as text

The column is `character` because two cells contain `"<1e-14"`, a
left-censored value someone entered by hand. One cell is enough to change the
type of the whole column.

Why this is an `error` and not a warning:

```{r}
as.numeric("<1e-14")
```

Silent `NA`. Those respondents would simply drop out of every biodiversity
metric, and nothing would say so. Worse is what happens if the column arrives
as a **factor** instead:

```{r}
as.numeric(factor(c("3.7e-13", "5.2e-13", "1.1e-13")))
```

Those are level codes, not values. `3.7e-13` becomes `1`. That is a
plausible-looking number with no warning at all, and every downstream
calculation would run happily and be completely wrong. This is the single most
destructive failure mode in the list, which is why `reiof` refuses to coerce
anything for you.

Note also that the magnitude and precision checks for `biodiv` do not appear in
the report above. Once a column fails the type check there is nothing numeric to
measure, so those checks are skipped rather than run on garbage. Fix the type,
re-run, and they appear.

**The fix** is a decision, not a conversion. What does `<1e-14` mean? If it is a
detection limit, you choose how to treat it; here we set it to missing and say
so.

```{r}
fixed <- broken
censored <- grepl("<", fixed$biodiv, fixed = TRUE)
sum(censored)
fixed$biodiv[censored] <- NA
fixed$biodiv <- as.numeric(fixed$biodiv)
```

```{r}
eiof_check_impact(fixed$biodiv, name = "biodiv", indicator = "biodiv")
```

## Fault 2 — greenhouse gases in the wrong unit

```{r}
subset(report, variable == "ghg" & check == "magnitude")$message
```

The check compares the median of your data against the published global 50th
percentile for that indicator. A ratio near 1000 is named as a likely
gram/kilogram error rather than left for you to spot.

Note that it reports an *unweighted* expectation here. Normally the expected
median is weighted to your sample's demographic mix, but the sex variable is
broken (fault 3), so no demographic group could be built. The check falls back
to an unweighted average rather than skipping, because a unit error and a sex
coding error are independent faults — you should not have to fix one to see the
other, and a factor of 1000 dwarfs any demographic weighting anyway.

This matters because the metric would otherwise run perfectly and report that
not one diet in the sample meets the greenhouse gas threshold — a result that
looks like a finding:

```{r}
grp_tmp <- eiof_age_sex_group(broken$age, broken$sex, male = 1, female = 0)
table(eiof_absolute_threshold(broken$ghg, grp_tmp, "ghg"), useNA = "ifany")
```

Zero diets below the threshold. Nothing errored. You could put that in a report.

**The fix:**

```{r}
fixed$ghg <- fixed$ghg / 1000
eiof_check_magnitude(fixed$ghg, "ghg", grp_tmp, name = "ghg")
```

## Fault 3 — sex coded 0/1 with no labels

```{r, error = TRUE}
eiof_age_sex_group(broken$age, broken$sex)
```

`reiof` will not guess. The reason is visible if you try it both ways:

```{r}
a <- eiof_age_sex_group(broken$age, broken$sex, male = 1, female = 0)
b <- eiof_age_sex_group(broken$age, broken$sex, male = 0, female = 1)
table(guess_A = factor(a, 1:8), guess_B = factor(b, 1:8))[1:4, 5:8]
```

The two readings are exact mirror images. Had the package picked one and been
wrong, every result would have been computed for the opposite sex, with no
error and nothing visibly odd in the output.

**The fix** — convert once, and the ambiguity is gone from the rest of the
script:

```{r}
fixed$sex <- eiof_as_sex_factor(broken$sex, male = 1, female = 0)
levels(fixed$sex)
grp <- eiof_age_sex_group(fixed$age, fixed$sex)
```

## Fault 4 — water rounded to four significant digits

```{r}
subset(report, variable == "water" & check == "precision")$message
```

This is what an Excel round trip does: the file displays `604.1` and saves what
it displays.

Does it matter? That is an empirical question, and the proximity check answers
it rather than leaving you to worry:

```{r}
eiof_check_precision(broken$water, grp, "water", name = "water")
```

Here, no observation sits close enough to the water threshold for the rounding
to change its classification — so for *this* analysis the lost digits are
harmless. For a different indicator or a different sample the answer could be
different, and you would get the row numbers.

There is no fix in code. Digits that were never read cannot be recovered; you
would have to go back to the source export. Worth knowing is not the same as
worth panicking about:

```{r}
nsig <- function(v) { for (d in 1:17) if (signif(v, d) == v) return(d); 17L }
c(broken = max(vapply(broken$water[1:50], nsig, integer(1))),
  demo   = max(vapply(eiof_demo$water[1:50], nsig, integer(1))))
```

## Re-run the checks

```{r}
report2 <- eiof_check_data(
  fixed,
  impacts = c(biodiv = "biodiv", ghg = "ghg", water = "water"),
  age = "age", sex = "sex", gdqs = "gdqs"
)
report2
```

No blocking problems. The remaining items are notes and the water precision
warning, which we have looked at and accepted.

```{r}
summary(report2)
```

## Now compute the metrics

```{r}
fixed$ghg_abs    <- eiof_absolute_threshold(fixed$ghg, grp, "ghg")
fixed$biodiv_abs <- eiof_absolute_threshold(fixed$biodiv, grp, "biodiv")
fixed$biodiv_bnd <- eiof_relative_band(fixed$biodiv, grp, "biodiv")
fixed$hesd       <- eiof_hesd(fixed$biodiv, fixed$gdqs, grp, form = "binary")
fixed$hdlei      <- eiof_hdlei(fixed$biodiv, fixed$gdqs, grp, form = "binary")

round(100 * c(
  ghg_below_threshold    = mean(fixed$ghg_abs,    na.rm = TRUE),
  biodiv_below_threshold = mean(fixed$biodiv_abs, na.rm = TRUE),
  hesd                   = mean(fixed$hesd,       na.rm = TRUE),
  hdlei                  = mean(fixed$hdlei,      na.rm = TRUE)
), 1)
```

Compare the greenhouse gas figure with the 0% we got before fixing the units.

## A reusable pattern

```r
# 1. check
report <- eiof_check_data(dat, impacts = c(...), age = "age", sex = "sex",
                          gdqs = "gdqs")
report

# 2. stop if anything is blocking
if (!summary(report)$pass) stop("fix the errors above before continuing")

# 3. compute
grp <- eiof_age_sex_group(dat$age, dat$sex)
dat$biodiv_abs <- eiof_absolute_threshold(dat$biodiv, grp, "biodiv")
# ...
```

Putting the `stop()` in writing is worth the two lines. A check you can skip is
a check that gets skipped on the day you are in a hurry.

## Error messages you may hit

| Message | Cause | Fix |
|---|---|---|
| `cannot tell which code means male` | numeric `sex` | `eiof_as_sex_factor(x, male =, female =)` |
| `these values of sex were not recognised` | an extra category, or a typo in a label | recode, or set `male`/`female` explicitly |
| `is a factor. as.numeric() on a factor returns level codes` | a factor impact column | `as.numeric(as.character(x))`, then re-check |
| `is character, not numeric` | a stray symbol, decimal comma or thousands separator | find the offending cells, decide what they mean |
| `column is integer` | impacts read as whole numbers | re-import as double; small values are already lost |
| `indicator must be one of` | a misspelled indicator | `eiof_list_indicators()` |
| `group has N value(s) outside 1-8` | `group` built by hand | use `eiof_age_sex_group()` |
| `percentile is required when metric = "rel"` | `eiof_lookup_cutoff()` without a percentile | add `percentile = 30` |

## What the checks do not do

They do not verify that your impact values are *correct* — only that they are
of a plausible type, sign, magnitude and precision. If the upstream matching of
food items to impact factors was wrong, every check here will pass and the
results will still be wrong. The checks narrow the space of silent failures;
they do not eliminate it.
